Towards a Verified Bzip2: A Functional Specification for Burrows-Wheeler

### 2019-05-03 draft_paper/draft_paper.tex
stash

The rest of this paper is organize as follows. \Cref{sec:bwt} details
the history of the BWT and its applications. \Cref{sec:ord} defines
what we need to know about a type $A$ to sort a $\var{list}~A$ as a
part of BWT. Following Bird et~al., \cref{sec:forwards_BWT} defines
the implementation of the forwards BWT in Coq, and
\cref{sec:inverse_BWT} derives an implementation of the inverse BWT
via program calculation. \Cref{sec:sorted_perm_stable} provides
multiple definitions of \var{Sorted}, \var{Permutation}, and
\var{StablePermutation}, and proves equivalence between them.
\Cref{sec:sort} implements insertion sort and radixsort, and proves
their correctness. \Cref{sec:BWT_proof} more formally proves the
theorems about the BWT mentioned in previous sections, including
\Cref{thm:unbwt_correct}.

@@

We can do this because there is only one way to sort the matrix, so
assuming \var{lexsort} and the radix-sort are correct, they must be
doing the same thing. We will prove this in
\cref{sec:sorted_perm_stable,sec:sort}.

%% In seeking to introduce an extra \verb|map rrot| term, we can note
%% that rotating and sorting an extra time doesn't change anything when
%% we are sorting the rotation matrix. This works because rotating the
%% last column to the front produces a matrix that is sorted on its last
%% \verb|n - 1| characters, so sorting stably on the first column will
%% re-sort the matrix. Crucially though, rotating the last column
%% forwards also permutes the entire matrix, so in addition to having a
%% sorted list, we have a sorted list of the same elements.

%% Using that we can show
%% \begin{equation*}
%%   lexsort (rots x)
%% = { because rrot permutes its elements }
%%   lexsort (rrot (rots x))
%% = { because hdsort permutes its elements }
%%   lexsort (hdsort (rrot (rots x)))
%% = { radix-sort }
%%   rep n (hdsort ∘ rrot) (hdsort (rrot (rots x)))
%% = { rep_absorb_r }
%%   rep (S n) (hdsort ∘ rrot) (rots x)
%% = { rep_split_l }
%%   hdsort (rrot (rep n (hdsort ∘ rrot) (rots x)))
%% = { radix-sort }
%%   hdsort (rrot (lexsort (rots x)))
%% \end{equation*}

%% From here we can calculate
%% \begin{equation*}
%%   cols (S j) (lexsort (rots l))
%% = cols (S j) (hdsort (map rrot (lexsort (rots l))))
%% = { hdsort commutes with cols (S j) }
%%   hdsort (cols (S j) (map rrot (lexsort (rots l))))
%% = { reduce cols (S j) }
%%   hdsort (prepend_col (map last (lexsort (rots l))) (cols j (lexsort (rots l)))
%% = { by IH and def of bwp }
%%   hdsort (prepend_col (bwp l) (recreate j (bwp l)))
%% = (fun x => hdsort (prepend_col x (recreate j x))) bwp
%% \end{equation*}

%% More precisely, we can characterize how \verb|map rrot| permutes
%% \verb|rots|
%% \begin{equation*}
%% map rrot (rots x) = rrot (rots x)
%% \end{equation*}

%% From here we can calculate
%% \begin{equation*}
%%   cols (S j) (lexsort (rots l))
%% = cols (S j) (hdsort (map rrot (lexsort (rots l))))
%% = { hdsort commutes with cols (S j) }
%%   hdsort (cols (S j) (map rrot (lexsort (rots l))))
%% = { reduce cols (S j) }
%%   hdsort (prepend_col (map last (lexsort (rots l))) (cols j (lexsort (rots l)))
%% = { by IH and def of bwp }
%%   hdsort (prepend_col (bwp l) (recreate j (bwp l)))
%% = (fun x => hdsort (prepend_col x (recreate j x))) bwp
%% \end{equation*}

%% We can use this to define \verb|recreate|:
%% \begin{equation*}
%% Fixpoint recreate (j : nat) (l : list A) : list (list A) :=
%%   match j with
%%   | O    => map (const []) l
%%   | S j' => hdsort (prepend_col l (recreate j' l))
%%   end.
%% \end{equation*}

@@

\subsection{\var{Permutation}}

The word \textit{permutation} has two related meanings that are often
used interchangably. We sometimes say that two lists are permutations of each
other if they contain all of the same elements, just appearing in a
different order. However, in mathematics we often say that a
permutation is actually the bijection on indices that defines the
correspondence between the two lists. For example, we say that $p$ is
the permutation that relates $l$ to $l'$ if for all indices $i$ and
$j$,
\begin{equation*}
  l_i = l'_{p(i)}
\end{equation*}
For consistency with Coq's built-in \var{Permuation} relation, in a
situation like this we will say that $l$ and $l'$ are permutations of
each other, and that $p$ is the \textit{permutation function} that
relates $l$ to $l'$. We say that two lists $l$ and $l'$ are
permutations of each other is precisely when
\vspace{-\parskip}
\begin{itemize}[nolistsep]
  \item $\var{length}~l = \var{length}~l' = n$
  \item there exists a permutation function $p : \mathbb{N}_{<n} \to
    \mathbb{N}_{<n}$, where $p$ is a bijection, such that for all indices $i$,
    \begin{equation*}
      l_i = l'_{p(i)}
    \end{equation*}
\end{itemize}

@@

%% In this section we prove the lemmas from the previous section:
%% \begin{enumerate}
%% \item
%% \begin{equation*}
%% map rrot (rots x) = rrot (rots x)
%% \end{equation*}

%% \item \verb|hdsort| commutes with \verb|cols (S j)|
%%   This proof is cool because in theory it can be proven via
%%   parametricity, but that's a metatheorem in Coq.

%% \item
%% \begin{equation*}
%% cols (S j) (map rrot l) = prepend_col (map last l) (cols j l)
%% \end{equation*}
%% \end{enumerate}

@@

\section{Proving BWT Correct}
\label{sec:BWT_proof}

@@

@@

@@

%% We prove this using a few lemmas about \var{findIndex},
%% \var{lexsort}, and \var{rots}.

%% \begin{lemma}[findIndex\_correct]
%%   For all elements $x$, lists $l$,
%%   \begin{equation*}
%%     (\exists y \in l,~x \equiv y) \implies l_{(\var{findIndex}~x~xs)} \equiv x
%%   \end{equation*}
%%   \begin{proof}
%%     This follows by induction on $xs$: At every step, either the
%%     current element is equivalent to $x$ and we return $0$, or there
%%     must be such a $y$ in the rest of the list.
%%   \end{proof}
%% \end{lemma}

%% \begin{lemma}[orig\_in\_sorted\_rots]
%%   For every non-empty list $l$,
%%   \begin{equation*}
%%     \exists y \in \var{lexsort}~(\var{rots}~l),~x = y
%%   \end{equation*}
%%   \begin{proof}
%%     Clearly $l$ is in $\var{rots}~l$ at index 0, and because
%%     \var{lexsort} produces a permutation of its input, $l$ must be
%%     in $\var{lexsort}~(\var{rots}~l)$ as well.
%%   \end{proof}
%% \end{lemma}

%% \begin{theorem}[bwn\_correct]
%%   For every non-empty list $xs$,
%%   \begin{equation*}
%%     (\var{lexsort}~(\var{rots}~xs))_{(\var{bwn}~xs)} = xs
%%   \end{equation*}
%%   \begin{proof}
%%     This follows from \textsf{findIndex\_correct} if we can show that
%%     $\exists y \in xs,~x = y$, which we have by \textsf{orig\_in\_sorted\_rots}.
%%   \end{proof}
%% \end{theorem}

>>>>

The rest of this paper is organize as follows. \Cref{sec:bwt} details
the history of the BWT and its applications. \Cref{sec:ord} defines
what we need to know about a type $A$ to sort a $\var{list}~A$ as a
part of BWT. Following Bird et~al., \cref{sec:forwards_BWT} defines
the implementation of the forwards BWT in Coq, and
\cref{sec:inverse_BWT} derives an implementation of the inverse BWT
via program calculation. \Cref{sec:sorted_perm_stable} provides
multiple definitions of \var{Sorted}, \var{Permutation}, and
\var{StablePermutation}, and proves equivalence between them.
\Cref{sec:sort} implements insertion sort and radixsort, and proves
their correctness. \Cref{appendix:BWT_proof} more formally proves the
theorems about the BWT mentioned in previous sections, including
\Cref{thm:unbwt_correct}.

@@

We can do this because there is only one way to sort the matrix, so
assuming \var{lexsort} and the radix-sort are correct, they must be
doing the same thing. We will prove this in
\cref{sec:unique,sec:sort}.

Using that we can show
\begin{align*}
     \var{lexsort}~(\var{rots}~x)
  &= \var{lexsort}~(\var{rrot}~(\var{rots}~x))
     && \text{because rrot permutes its elements} \\
  &= \var{lexsort}~(\var{hdsort}~(\var{rrot}~(\var{rots}~x)))
     && \text{because hdsort permutes its elements} \\
  &= (\var{hdsort} \comp \var{rrot})^n~(\var{hdsort}~(\var{rrot}~(\var{rots}~x)))
     && \text{radix-sort} \\
  &= (\var{hdsort} \comp \var{rrot})^{n+1}~(\var{rots}~x)
  && \text{rep\_absorb\_r} \\
  &= \var{hdsort}~(\var{rrot}~((\var{hdsort} \comp \var{rrot})^n~(\var{rots}~x)))
     && \text{ rep\_split\_l } \\
  &= \var{hdsort}~(\var{rrot}~(\var{lexsort}~(\var{rots}~x)))
     && \text{ radix-sort }
\end{align*}

In seeking to introduce an extra \var{map rrot} term, we can note that
rotating and sorting an extra time doesn't change anything when we are
sorting the rotation matrix. This works because rotating the last
column to the front produces a matrix that is sorted on its last $n -
1$ characters, so sorting stably on the first column will re-sort the
matrix. Crucially though, rotating the last column forwards also
permutes the entire matrix, so in addition to having a sorted list, we
have a sorted list of the same elements.

More precisely, we can characterize how $\var{map}~\var{rrot}$
permutes \var{rots}.
\begin{equation*}
  \var{map}~\var{rrot}~(\var{rots}~x) = \var{rrot}~(\var{rots}~x)
\end{equation*}

From here we can calculate
\begin{align*}
  &\var{cols}~(j+1)~(\var{lexsort}~(\var{rots}~l)) \\
&= \var{cols}~(j+1)~(\var{hdsort}~(\var{map}~\var{rrot}~(\var{lexsort}~(\var{rots}~l)))) \\
&= \var{hdsort}~(\var{cols}~(j+1)~(\var{map}~\var{rrot}~(\var{lexsort}~(\var{rots}~l))))
&&\text{ \var{hdsort} commutes with $\var{cols}~(j+1)$ } \\
&= \var{hdsort}~(\var{prepend\_col}~(\var{map}~\var{last}~(\var{lexsort}~(\var{rots}~l)))~(\var{cols}~j~(\var{lexsort}~(\var{rots}~l)))
  &&\text{ reduce $\var{cols}~(j+1)$ } \\
&= \var{hdsort}~(\var{prepend\_col}~(\var{bwp}~l)~(\var{recreate}~j~(\var{bwp}~l)))
  &&\text{ by IH and def of bwp } \\
&= (\lambda x.~\var{hdsort}~(\var{prepend\_col}~x~(\var{recreate}~j~x)))~\var{bwp}
\end{align*}

We can use this to define \var{recreate}:
\lstinputlisting[
  widthgobble=2, firstline=156, lastline=160,
]{../theories/BWT/BurrowsWheeler.v}

@@

\subsection{\var{Permutation}}
\label{subsec:perm}

The word \textit{permutation} has two related meanings that are often
used interchangably. We sometimes say that two lists are permutations of each
other if they contain all of the same elements, just appearing in a
different order. However, in mathematics we often say that a
permutation is actually the bijection on indices that defines the
correspondence between the two lists. For example, we say that $p$ is
the permutation that relates $l$ to $l'$ if for all indices $i$ and
$j$,
\begin{equation*}
  l_{p(i)} = l'_i
\end{equation*}
For consistency with Coq's built-in \var{Permuation} relation, in a
situation like this we will say that $l$ and $l'$ are permutations of
each other, and that $p$ is the \textit{permutation function} that
relates $l$ to $l'$. We say that two lists $l$ and $l'$ are
permutations of each other is precisely when
\vspace{-\parskip}
\begin{itemize}[nolistsep]
  \item $\var{length}~l = \var{length}~l' = n$
  \item there exists a permutation function $p : \mathbb{N}_{<n} \to
    \mathbb{N}_{<n}$, where $p$ is a bijection, such that for all indices $i$,
    \begin{equation*}
      l_{p(i)} = l'_i
    \end{equation*}
\end{itemize}

@@

\subsection{\var{StablePerm}}
\label{subsec:stable}

A stable sort is a sort that doesn't swap any two elements considered
equivalent by the ordering. This is relevant when the induced
equivalence relation does not imply equality, so that we can
distinguish between elements that the sort considers identical. A good
example of this is in radixsort: when we sort on a single column, the
sort can't tell the difference between strings that contain equal
characters in that column, but it is crucial that it respect the
initial ordering in cases of a tie.

For stable sorts, we need a stronger property than \var{Permutation}.
We can extend our mathematical definition. We say that two lists $l$
and $l'$ are stable permutations of each other when we have a
permutation $p$ that relates them, and that satisfies an additional
stability property: for all indices $i$ and $j$ where
$(\var{apply}~p~l)_i \equiv (\var{apply}~p~l)_j$
\begin{gather*}
  i < j \implies p(i) < p(j)
\end{gather*}
Equivalently, we can say that for all indices $i$ and $j$ where $l_j \equiv
l_j$ we must have
\begin{gather*}
  i < j \implies p^{-1}(i) < p^{-1}(j)
\end{gather*}

Based on this, we define a stronger version of the \var{PermFun}
property, \var{StablePermFun}. However, while $\var{PermFun}~n~p$
means that $p$ represents a permutation function on lists of length
$n$, we need to specify the list with resepct to which $p$ is stable,
so we say $\var{StablePermFun}~l~p$ for some list $l$.
\lstinputlisting[
  widthgobble=2, firstline=557, lastline=561,
]{../theories/BWT/Sorting/PermFun.v}
As above, we can also define this property using the inverse
permutation:
\lstinputlisting[
  widthgobble=2, firstline=563, lastline=567,
]{../theories/BWT/Sorting/PermFun.v}
As usual, we will prove these definitions equivalent. First, though,
we need to define \var{image} and \var{preimage}.
\lstinputlisting[
  widthgobble=2, firstline=85, lastline=86,
]{../theories/BWT/Sorting/PermFun.v}
We will not prove it here, but $\var{preimage}~p$ and $\var{image}~p$
are both of bijections of type $\mathbb{N}_{<n} \to \mathbb{N}_{<n}$
and are invereses of each other. We also have the following two
theorems:
\begin{theorem}[nth\_preimage\_apply, nth\_image\_apply]
  For all lists $l$, permutations $p$, and indices $i$,
  \begin{align*}
    (\var{apply}~p~l)_{p^{-1}(i)} = l_i && l_{p(i)} = (\var{apply}~p~l)_i
  \end{align*}
\end{theorem}
\begin{proof}
  The first equation (\var{nth\_preimage\_apply}) holds because:
  \begin{align*}
       (\var{apply}~p~l)_{p^{-1}(i)}
    &= (\var{map}~(\lambda j.~l_j)~p)_{\var{findIndex}~i~p} \\
    &= l_{p_{(\var{findIndex}~i~p)}} \\
    &= l_i
  \end{align*}
  and the second follows from this:
  \begin{equation*}
      (\var{apply}~p~l)_i
    = (\var{apply}~p~l)_{p^{-1}(p(i))}
    = l_{p(i)}
    \qedhere
  \end{equation*}
\end{proof}

\begin{theorem}[StablePermFun\_iff]
  For all lists $l$ and permutation $p$,
  \begin{equation*}
    \var{StablePermFun}~l~p \iff \var{StablePermFun\_preimage}~l~p
  \end{equation*}
\end{theorem}
\begin{proof}
  We will only prove the $\implies$ direction; the other direction is
  very similar. We have that $l_i \equiv l_j$, and that $i < j$, and we
  have to show that $p^{-1}(i) < p^{-1}(j)$. We will show this by
  proving that $p^{-1}(i) \ge p^{-1}(j)$ leads to a contradiction.

  Suppose $p^{-1}(i) \ge p^{-1}(j)$. We actually know that $p^{-1}(i) >
  p^{-1}(j)$ because $i \neq j$. From $l_i \equiv l_j$ and
  $\var{StablePermFun}~l~p$, along with $p^{-1}(j) < p^{-1}(i)$, we
  know that $p(p^{-1}(j)) < p(p^{-1}(i))$. However, this implies $j <
  i$, which is a contradiction. Therefore, $p^{-1}(i) < p^{-1}(j)$.
\end{proof}

Now, we can define \var{StablePermEx} to describe when two lists are
stable permutations of each other:
\lstinputlisting[
  widthgobble=2, firstline=607, lastline=608,
]{../theories/BWT/Sorting/PermFun.v}

While \var{StablePermEx} corresponds to \var{PermutationEx}, how can
we modify \var{Permutation} to create an inductive relation for stable
lists? Intuitively, the problematic inference rule is
\var{Permutation\_swap}: we can only swap two elements and preserve
stability if they are not equivalent. With this fix, we have \var{StablePermInd}:
\begin{gather*}
  \infer[\var{StablePermInd\_nil}]{\var{StablePermInd}~[]~[]}{}
  \\ \\
  \infer[\var{StablePermInd\_skip}]{\var{StablePermInd}~(x :: l)~(x :: l')}{%
    \var{StablePermInd}~l~l'
  }
  \\ \\
  \infer[\var{StablePermInd\_swap}]
        {\var{StablePermInd}~(x :: y :: l)~(y :: x :: l)}
        {x \not\equiv y}
  \\ \\
  \infer[\var{StablePermInd\_trans}]{\var{StablePermInd}~l~l''}{
    \var{StablePermInd}~l~l' & \var{StablePermInd}~l'~l''
  }
\end{gather*}

Unforunately, neither of these two definitions will be particularly
useful for proving the theorems we would like to prove with them. The
best definition for stability does not come by analogy with
\var{Permutation}. The intuition is that for any two lists that are
stable permutations of each other, if we filter each list by
equivalence with any element, the resulting lists should be identical.
This is because they must have the same elements, since the original
lists are permutations of one another, and those elements must be in
the same order, since the original lists are specifically stable
permutations of one another.
\lstinputlisting[
  widthgobble=2, firstline=21, lastline=22,
]{../theories/BWT/Sorting/Stable.v}

We will prove equivalence between these three definitions, but first
we will mention some properties of \var{Stable}. \var{Stable} is an
equivalence relation, meaning it is reflexive, transitive, and symmetric.

\begin{lemma}[StablePerm\_app]
  For all lists $l$, $l'$, $m$ and $m'$,
  \begin{equation*}
    \var{StablePerm}~l~l' \implies \var{StablePerm}~m~m' \implies
    \var{StablePerm}~(l \dplus m)~(l' \dplus m')
  \end{equation*}
\end{lemma}
\begin{proof}
  This follows from the distributivity of \var{filter} over \var{app}.
\end{proof}

\begin{lemma}[StablePerm\_skip]
  For all lists $l$ and $l'$, and elements $a$,
  \begin{equation*}
    \var{StablePerm}~l~l' \implies \var{StablePerm}~(a::l)~(a::l')
  \end{equation*}
\end{lemma}
\begin{proof}
  For any $x$, we have to show $\var{filter}~(\equiv x)~(a::l) =
  \var{filter}~(\equiv x)~(a::l')$. If $x \equiv a$, $a$ is added to both lists,
  and if $x \not\equiv a$ it is added to neither.
\end{proof}

\begin{lemma}[StablePerm\_swap]
  For all lists $l$ and $l'$, and elements $a$ and $b$ where $a \not\equiv b$,
  \begin{equation*}
     \var{StablePerm}~(a::b::l)~(b::a::l)
  \end{equation*}
\end{lemma}
\begin{proof}
  For any $x$, we have to show $\var{filter}~(\equiv x)~(a::b::l) =
  \var{filter}~(\equiv x)~(b::a::l)$. The proof follows from case analysis
  on $x \equiv a$ and $x \equiv b$.
\end{proof}

\begin{lemma}[StablePerm\_cons\_app]
  For all lists $l$, $l_1$ and $l_2$, and elements $a$ such that $\forall b
  \in l_1,~a \not\equiv b$,
  \begin{equation*}
     \var{StablePerm}~l~(l_1 \dplus l_2) \implies \var{StablePerm}~(a
     :: l)~(l_1 \dplus a :: l_2)
  \end{equation*}
\end{lemma}
\begin{proof}
  For any $x$, we have to show
  \begin{equation*}
    \var{filter}~(\equiv x)~(a::l) = \var{filter}~(\equiv x)~(l_1 \dplus a :: l_2)
  \end{equation*}
  If $x \not\equiv a$ the proof is easy, so assume $x \equiv a$. Then we have
  \begin{align*}
      \var{filter}~(\equiv x)~(l_1 \dplus a :: l_2)
    &= \var{filter}~(\equiv x)~l_1 \dplus \var{filter}~(a :: l_2) \\
    &= [] \dplus a::\var{filter}~l_2 \\
    &= a::\var{filter}~l
  \end{align*}
  where we know that $\var{filter}~(\equiv x)~l_1 = []$ because $\forall b \in
  l_1,~a \not\equiv b$, and the last step follows from
  $\var{StablePerm}~l~(l_1 \dplus l_2)$.
\end{proof}

\begin{theorem}[StablePerm\_destr]
  For all lists $l$ and $l'$, and elements $h$ and $h'$ such that $h
  \not\equiv h'$,
  \begin{gather*}
    \var{StablePerm}~(h :: l)~(h' :: l') \implies \\
    \exists~l_1~l_2, ~l = l_1 \dplus h' :: l_2 \land (\forall x \in l_1, h' \not\equiv x)
  \end{gather*}
\end{theorem}
\begin{proof}
  Let $l_1 = \var{take\_while}~(\not\equiv h')~l$ and $l_2 =
  \var{drop\_while}~(\not\equiv h')~l$. Clearly $l = l_1 \dplus l_2$, and
  because $\var{StablePerm}~(h :: l)~(h' :: l')$, we know the first
  element of $l_2$ must be $h'$. Let $l_2'$ be the rest of $l_2$, and
  we have two lists $l_1$ and $l_2'$ that satisfy the theorem.
\end{proof}

\begin{theorem}[StablePerm\_length]
  For all lists $l$ and $l'$
  \begin{equation*}
    \var{StablePerm}~l~l' \implies \var{length}~l = \var{length}~l'
  \end{equation*}
\end{theorem}
\begin{proof}
  We reason by induction on $n = \var{length}~l$. In the inductive
  case, we need to show that $\var{length}~(h::t) =
  \var{length}~(h'::t')$ given that $\var{StablePerm}~(h::t)~(h'::t')$
  and the induction hypothesis that for all lists of length $n$,
  \var{StablePerm} implies their lengths are equal.

  If $h \equiv h'$, then from $\var{StablePerm}~(h::t)~(h'::t')$ we know
  that $h = h'$. We can then remove $h$ from the front of $h::t$ and
  $h::t'$ and apply the induction hypothesis.

  If $h \not\equiv h'$, we can destructure $t$ and $t'$ as $l_1 \dplus h'
  :: l_2$ and $l_1' \dplus h :: l_2'$ respectively, using
  \var{StablePerm\_destr}. Then,
  \begin{equation*}
      \var{length}~t
    = \var{length}~(h' :: l_1  \dplus l_2)
    = \var{length}~(h  :: l_1' \dplus l_2')
    =\var{length}~t'
  \end{equation*}
  where the first and last steps follow from the induction hypothesis,
  using the fact that $\var{StablePerm}~t~(h'::l_1 \dplus l_2)$ and
  $\var{StablePerm}~t'~(h::l_1' \dplus l_2')$.
\end{proof}

\begin{theorem}[StablePerm\_Perm]
  For all lists $l$ and $l'$,
  \begin{equation*}
    \var{StablePerm}~l~l' \implies \var{Permutation}~l~l'
  \end{equation*}
\end{theorem}
\begin{proof}
  We reason by strong induction on the length of both lists (we know
  the lengths are equal by \var{StablePerm\_length}). In the inductive
  case, we need to show $\var{Permutation}~(h :: t)~(h' :: t')$. If $h
  \equiv h$, we know that $h = h'$, and the proof follows by
  \var{perm\_skip}.

  If $h \not\equiv$, we again destructure $t$ and $t'$ as $l_1 \dplus h' ::
  l_2$ and $l_1' \dplus h :: l_2'$ respectively, using
  \var{StablePerm\_destr}.
  \begin{align*}
    h::t &= h :: l_1 \dplus h' :: l_2 \\
    &\eqperm h :: h' :: l_1 \dplus l_2 \\
    &\eqperm h' :: h :: l_1 \dplus l_2 \\
    &\eqperm h' :: h :: l_1' \dplus l_2' \\
    &\eqperm h' :: l_1' \dplus h :: l_2' = h'::t
  \end{align*}
  The crucial step is $l_1 \dplus l_2 \eqperm l_1' \dplus l_2'$, which
  follows from the induction hypothesis.
\end{proof}

\begin{theorem}[StablePermInd\_iff]
  For all lists $l$ and $l'$
  \begin{equation*}
    \var{StablePerm}~l~l' \iff \var{StablePermInd}~l~l'
  \end{equation*}
\end{theorem}
\begin{proof}
  The ($\impliedby$) direction follows easily by induction on
  $\var{StablePermInd}~l~l'$, using \var{StablePerm\_skip},
  \var{StablePerm\_swap}, and \var{StablePerm\_trans}.

  The ($\implies$) direction is proved very similarly to
  \var{StablePerm\_Perm}, using \var{StablePerm\_destr} to split up
  the lists.
\end{proof}

Now we prove equivalence with \var{StablePermEx}.

\begin{theorem}[apply\_correct\_stable]
  For all lists $l$ and permutation functions $p$,
  \begin{equation*}
    \var{StablePermFun}~l~p \implies \var{StablePerm}~(\var{apply}~p~l)~l
  \end{equation*}
\end{theorem}
\begin{proof}
  We reason by induction on the length of $l$. In the inductive case,
  we need to show
  \begin{equation*}
    \var{StablePerm}~(\var{apply}~(i :: p)~(a :: l))~(a :: l)
  \end{equation*}

  Using $\eqstable$ to chain \var{StablePerm}s transitively,
  \begin{align*}
    \var{apply}~(i :: p)~(a :: l)
    &= (a :: l)_i  ::
    \var{apply}~(\var{rem\_PermFun}~i~p)~(\var{rem\_nth}~i~(a :: l)) \\
    &\eqstable (a :: l)_i :: \var{rem\_nth}~i~(a :: l) \\
    &\eqstable (a :: l)_i
  \end{align*}
  where we rely on the induction hypothesis between the first and
  second lines.
\end{proof}

\begin{theorem}[StablePermEx\_iff]
  For all lists $l$ and $l'$
  \begin{equation*}
    \var{StablePerm}~l~l' \iff \var{StablePermEx}~l~l'
  \end{equation*}
\end{theorem}
\begin{proof}
  We use \var{StablePermInd\_iff} and prove $\var{StablePermInd}~l~l'
  \iff \var{StablePermEx}~l~l'$.

  The $\implies$ direction is very similar to the $\implies$ direction
  of the \var{PermutationEx\_iff} proof: we reason by induction on
  $\var{StablePermInd}~l~l'$, constructing the same permutation
  functions in each case, but this time we have to prove that it is
  stable as well. The $\impliedby$ direction follows from
  \var{apply\_correct\_stable}.
\end{proof}

\section{Uniqueness of Stable Sorts}
\label{sec:unique}

Using our definitions of \var{StablePerm} and \var{Sorted}, we can
prove the theorem that allowed us to replace \var{lexsort} with
radixsort.

\begin{theorem}[stable\_sort\_unique]
  \begin{equation*}
    \var{Sorted}~l \implies \var{Sorted}~l' \implies
    \var{StablePerm}~l~l' \implies l = l'.
  \end{equation*}
\end{theorem}

With \var{lexsort}, we are guaranteed stability because we are using
an order, not just a preorder, and for orders all sorts are stable.

\begin{theorem}[all\_perm\_stable]
  If we have that equivalence implies Leibniz equality, then for all
  lists $l$ and $l'$
  \begin{equation*}
      \var{Permutation}~l~l' \implies \var{StablePerm}~l~l'
  \end{equation*}
\end{theorem}

@@

\lstinputlisting[
  widthgobble=2, firstline=11, lastline=21,
]{../theories/BWT/Sorting/InsertionSort.v}

\begin{theorem}[sort\_sorted]
  \begin{equation*}
    \var{Sorted}~(\var{sort}~l)
  \end{equation*}
\end{theorem}

\begin{theorem}[sort\_stable]
  \begin{equation*}
    \var{Stable}~(\var{sort}~l)
  \end{equation*}
\end{theorem}

\lstinputlisting[
  widthgobble=2, firstline=22, lastline=23,
]{../theories/BWT/Sorting/RadixSort.v}

\begin{theorem}[radixsort\_sorted]
  \begin{equation*}
    (\forall x \in l,~\var{length}~x = n) \implies \var{Sorted}~(\var{radixsort}~l~n)
  \end{equation*}
\end{theorem}

@@

\section{Proving BWT Correct}
\label{appendix:BWT_proof}

In this section we prove the lemmas from the derivation that we left out:
\begin{enumerate}
\item
\begin{equation*}
  \var{map}~\var{rrot}~(\var{rots}~x) = \var{rrot}~(\var{rots}~x)
\end{equation*}

\item \var{hdsort} commutes with $\var{cols} (j+1)$.
  This proof is cool because in theory it can be proven via
  parametricity, but that's a metatheorem in Coq.

\item
  \begin{equation*}
    \var{cols}~(j+1)~(\var{map}~\var{rrot}~l) = \var{prepend\_col}~(\var{map}~\var{last}~l)~(\var{cols}~j~l)
  \end{equation*}
\end{enumerate}

@@

\begin{theorem}[compose\_apply]
  For all lists $l$ and permutation functions $p_1$ and $p_2$
  \begin{equation*}
      \var{apply}~(p_2 \comp p_1)~l = \var{apply}~p2~(\var{apply}~p1~l)
  \end{equation*}
\end{theorem}

@@
